### 2011-12-09 paper.tex
added the tex files

\documentclass[10pt]{article}
\pagestyle{plain}

\usepackage[margin=2.5cm]{geometry}
\usepackage{multicol}
\usepackage{url}

\begin{document}

\title{CMSC818s: Project 3 \\
A Brief Survey of Modern Virtualization Research}
\author{}
\date{}

\maketitle

\vspace{-2cm}

\begin{center}
\begin{multicols}{2}
  Greg Benjamin \\
  gregben@cs.umd.edu \\
  Austin Myers \\
  amyers@cs.umd.edu
\end{multicols}
Department of Computer Science \\
University of Maryland, College Park
\end{center}

\begin{multicols*}{2}

\section{Introduction}
\label{sec:intro}

The concept of virtualization has been around for a long time, reaching back
to the 1960s and the introduction of virtual memory in timeshared machines.
However, in the last decade or so it's become a very exciting area of research,
particularly motivated by a new market for server virtualization.  Commerical
products like VMWare and open-source platforms like Xen \cite{ref:xen} have made
it possible for companies to host dozens of instances of their servers
on the same hardware.  Companies can now spend less on metal, can seamlessly and
easily migrate servers from one machine to another, and can even vary their fleet
size on the fly in response to fluctuations in server traffic by spinning up new
instances.  In addition to the commercial successes that have come as a result
of server virtualization, there have been tremendous breakthroughs in the research
community as well, ranging from new solutions to problems of isolation and protection
to novel new algorithms for scheduling and resource allocation.

Despite the flurry of interest in this area, there are still many interesting
and difficult problems to be solved.  Sharing finite resources
fairly and safely amongst many competing processes has never been easy, but
the popularity of virtualization in the cloud makes scalability and efficiency
bigger concerns than ever before.  Furthermore, the commercialization of
virtual machines raises questions of how to ensure quality of service in
an environment where everyone is in contention for the hardware.  In this
paper, we will examine some of the recent literature in this area and
seek to draw some conclusions about the problem space and the directions
of future work.

The rest of the paper is structured as follows: in Section \ref{sec:background}
we provide background information on the history of virtualization and the
sorts of techniques which are utilized today.  Section \ref{sec:summaries}
presents summaries of three recent virtualization papers from the
9\textsuperscript{th} USENIX Symposium on Operating Systems Design
and Implementation, as well as a fourth from (TBD!).  Section
\ref{sec:relationships} analyses these papers in terms of how they
relate to each other and how they approach the larger problem space.
Finally, in Section \ref{sec:conclusion} we conclude by discussing our
thoughts and opinions on the problem space as a whole, and postulate
about how virtualization may evolve in the future.

\section{Background}
\label{sec:background}

The concept of virtualization reaches back to the 1960s, when developers
of large time-sharing systems sought ways to give each user the illusion that
they had sole access to a private, virtual computer, despite the fact that
many users were in fact sharing computing resources.  Once private physical
computers became available, popular, and cheap in the 80s, virtualization
took a backseat to other areas of research.  However, many of the ideas
put forth in these early days lived on in work involving virtual memory
and operating system protection paradigms.

Server virtualization arrived on the scene about a decade ago, with VMWare
appearing in 1999 and Xen in 2003.  Since then, research and commercial interest
in virtualization has sparked significantly, especially in conjunction
with the rise of cloud computing and infrastructure-as-a-service.

In a virtualized context, we have many \emph{virtual machines}(VM) sharing the same
physical hardware.  In a \emph{fully virtualized} system, each VM has the illusion
that it alone has sole access to the hardware, although in a \emph{paravirtualized}
VMs may be modified to be aware of the fact that they are virtualized.  This
sharing of resources is possible because access to the hardware is regulated by
a special piece of software, the \emph{virtual machine monitor} or \emph{hypervisor}.
In many ways, the hypervisor plays the role of the kernel in a traditional operating
system - it contains the privileged software that can directly access the hardware.

The hypervisor (and potentially other administrative software) runs in the \emph{host}
domain (in Xen, this is called \emph{dom0}), while other OSes running in VMs are 
referred to as \emph{guests} (in Xen, all guests run in \emph{domU}).  Generally, it is
crucial that all guests are isolated from one another, so that the actions of one guest
cannot adversely affect the performance of any other.  This is a difficult property
to provide in a system where all guests are in contention for hardware resources.

In the early days of virtualization, there was little-to-no hardware support,
and so the multiplexing of VMs onto the physical hardware was accomplished by
\emph{emulating} hardware devices in the software for each VM.  This had the unfortunate
drawback of being extremely slow and inefficient, because the hardware was emulated by
the hypervisor, and so most device accesses by a guest VM resulted in multiple traps
into the hypervisor.  In the past decade or so, chipsets have become more friendly,
and most architectures (notably the x86) now offer hardware-level support for
virtualization, along with a set of privileged instructions for making use of it.
This has allowed hardware virtualization to become fast enough to be viable,
so that virtualized systems are now reasonably competitive with their native
counterparts.

\section{Paper Summaries}
\label{sec:summaries}

\subsection{The Turtles Project: Design and Implementation of Nested Virtualization}
\label{sec:summaries/turtles}

The Turtles Project \cite{ref:turtles} aims to tackle the problem of
nested virtualization; rather than being able to run multiple guest OSes
on a single host machine, the authors want to take things a step farther
and be able to run multiple guest hypervisors.  This is actually a
much harder problem than its single-level counterpart, because while
most hardware now supports virtualization, chips are generally designed
to only handle a single level.  Furthermore, the strawman solution of
using software to emulate the virtualization-enabled hardware for the next level is
extremely expensive, because any kernel traps generated at the highest
level (from memory paging or device I/O) generate traps in the levels
below as well, resulting in cascading crossings of protection
boundaries and a serious performance penalty.

Key contributions of this work include the design and implementation
of a truly efficient nested VM for the x86 architecture, including novel
techniques for \emph{multidimensional paging} and
\emph{multi-level device assignment}.  The Turtles hypervisor is able to
run unmodified guest hypervisors and OSes with a performance penalty of
as low as 6-8\%.  The paper also presents what the authors claim is the
first thorough analysis and evaluation of nested hypervisor performance,
with an eye to the identification and reduction of performance bottlenecks.

Turtles is designed for the x86 architecture, which subscribes to a
single-level hardware support model.  This means that there are only two
levels of execution privilege.  Turtles runs its host hypervisor, which
we will call $L_0$, at the most privileged level of execution.  Now, suppose
we have a hypervisor running above $L_0$ at level $L_1$, and that hypervisor
is running a guest OS at level $L_2$.  Everything running at $L_1$ and $L_2$
has to run at the same privilege level, and only $L_0$ is privileged enough
to interact with the hardware.  Essentially, levels $L_1$ through $L_n$ only
exist logically, and in reality they are all multiplexed to be running directly
on top of $L_0$.  All traps generated at $L_2$ go immediately to $L_0$, which
may handle them itself, or forward them back to $L_1$ for handling if hardware
interaction is not required.

There are two difficult problems here that turtles provides a unique solution for.
The first involves page tables for the MMU.  In a nested model, virtual addresses for $L_2$
will need to be translated three times (first to physical addresses at $L_2$, then to $L_1$
addresses, and finally to $L_0$ addresses)
before they are readable by the hardware.  Modern MMUs provide support for virtualization
in the form of two distinct translation tables to make this process fast, but this is not
sufficient for more than one level of virtualization.  Furthermore, providing these
translation tables by emulating them in software is slow, particularly on updates,
for the reasons described above.  The paper's novel solution here is to use
\emph{multidimensional page tables}.  Essentially, each level keeps its own
shadowed page tables in software, but they are read-only.  On a page fault in $L_2$,
$L_0$ catches the trap and forwards it to $L_1$.  If $L_1$ also faults, $L_0$ allows
it to update its shadow table with the $L_2$-to-$L_1$ translation, but then applies
its tranlation for $L_1$ addresses to update the hardware table with the correct
$L_2$-to-$L_0$ translation before returning.  In this way, the hardware table at $L_0$
contains all the shadow tables above it compressed into a single table, so that
any MMU accesses can look up the correct translations directly from the hardware
without having to go through cascading levels of shadow tables first.  Turtles
also employs ``huge pages'' and Virtual Processor ID tags on page table entries
to ensure that page tables remain small and fast.

The second difficulty deals with device I/0.  Traditional approaches to this problem
involve either software emulation (which is slow), installing modified drivers in the
guest (which is not preferrable), or assigning devices to guests as they have need
of them (which is hard in the nested case).  Turtles uses something called
\emph{multi-level device assignment}, which is essentially an adaptation of the
multidimensional page table idea to the device I/O problem.  Modern chipsets
include a component called an IOMMU, which resides between the DMA and main memory
and performs translations just like the MMU does for the CPU.  By shadowing read-only
IOMMU translation tables at higher levels, and then compressing them into a single
hardware-level table, the DMA can avoid the same cascading translation issue that
would affect the MMU, thus making DMA I/O access very efficient.  For Memory-Mapped I/O
and Network Port I/O, address translation is not an issue, and so here techniques for
single-level device assignment are used without any penalty.

The Turtles paper provides an extensive evaluation section, with thorough experimentation
comparing performance of applications running on the native OS, a single-level VM, and the
Turtles two-level nested VM.  The authors measure and discuss performance on microbenchmarks,
I/O intensive workloads, and workloads that are designed to fault to the kernel constantly.
The paper also analyzes the success of multidimensional paging and multi-level device
assignment by testing these techniques against more traditional ways of handling these
problems.  In general, the Turtles hypervisor far outperforms any other approach to
nested virtualization considered, and is able to minimize the overhead of a second
level of virtualization to as low as 6-8\%, which is rather impressive.  No experiments
were conducted involving more than two nested levels of virtualization, although analysis
seemed to indicate that most of the additional overhead incurred by adding a second level
came from cache pollution and the overuse of privileged instructions where it was not
necessary.

\subsection{mClock: Handling Throughput Variability for Hypervisor IO Scheduling}
\label{sec:summaries/mclock}

The mClock paper \cite{ref:mclock} is not targeted at nested virtualization,
but further explores
the issue of fair I/O device assignment raised in the Turtles paper.  This turns
out to be a very difficult problem even in the single-level case, particularly
when applied to virtualization in the cloud, where networked devices are
often shared not just between VMs running on one host, but between many hosts
across the network.  In the interest of fairness, a good hypervisor will try to
share device throughput equally across its guests, but in the networked case, the
total amount of capacity to share may fluctuate due to traffic from other
networked clients.  Furthermore, many OSes run their own device access algorithms
above the hypervisor level to improve performance; for example, physical disk
accesses may be reordered by the kernel to take advantage of spatial locality
when moving the disk head.  These sorts of optimizations are often difficult
for the hypervisor to multiplex without loss of efficiency.

The mClock algorithm seeks to fairly schedule the device I/O of multiple guests,
while maintaining each guest's limit (maximum throughput requirement), reservation
(minimum requirement), and weight (proportional share of capacity relative to
other guests).  The algorithm uses a ``novel, lightweight'' tagging scheme to
achieve this despite fluctuation in total capacity.  Previous attempts to solve
this problem were not able to achieve both of these goals simultaneously.

The mClock algorithm works by assigning numerical tags to each queued request.
Essentially, tags for VM $i$ with proportional weight $w_i$ will be spaced by $1/w_i$.
Requests will be executed in the order of their assigned tags, which ensures that
VM $i$ receieves his proportional share.  This approach requires a global clock
for tag assignment, so that idle clients remain synced with active ones and do not
amass ``idle credit'', thus gaining preferential treatment.  The tag for VM $i$ and
request $j$ is set as follows:

\[
	Tag(i, j) = max(curr~time, Tag(i, j-1) + 1/w_i)
\]

After the tagging, the request with the lowest tag will be scheduled.

In actuality, things are a little bit more complicated.  mClock maintains
3 types of tags and 3 global clocks, one each for reservations, weights,
and limits.  Each iteration of the algorithm is deemed to be
either a ``constraint-based phase'', in which some VM has not received
capacity above its reservation, or a ``weight-based phase'', in which the
next VM to be scheduled will be chosen by proportional share.  This is easily
determined after the tagging:  if any VM $i$ has request $j$ such that the
reservation tag $R(i, j)$ is less than the current time, than VM $i$ is currently
receiving less capacity than its reservation, and so the current phase is
constraint based.  Otherwise, the phase is weight-based, and so the algorithm
just picks a request with minimum weight tag $P(i, j)$ whose limit tag
$L(i, j)$ is also less than the current time (otherwise, VM $i$ is receiving
more capacity than its limit, and so it should not be scheduled).

The paper describes some further tweaks to the algorithm, such as allowing
a limited amount of idle credit to be saved up by idle VMs, or breaking ``large''
requests into a series of smaller requests to account for request latency
in the scheduler.  There's also a modifed dmClock algorithm for use with
a distributed storage system, which multiplexes VM requests fairly across multiple
storate servers.  Finally, there's a heuristic for setting reservations for
all VMs by default, so that no VM will be starved.

The authors include a lengthy evaluation of the performance of their algorithm.
They discuss many experiments involving different workloads and different combinations
of weights, limits, and reservations, and even test multiple types of devices
being assigned to analyze the effects of device latency and varying throughput.
Overall, mClock is extremely effective at meeting its guarantees for weights, limits,
and reservations when there is enough capacity to satisfy all requests, and
the algorithm degrades gracefully by sharing capacity proportionally when
the total capacity is insufficient.

The algorithm is successful at delivering on its guarantees despite contention,
which is important for ensuring VM isolation, and also has implications in the cloud,
where meeting quality of service guarantees determine financial success.

\subsection{Virtualize Everything but Time}
\label{sec:summaries/time}

The ``Virtualize Everything but Time'' paper \cite{ref:time} notes that accurate
timekeeping has become very important in today's computing systems, with
applications to everything from systems measurement and distributed database
consistency to high-speed trading and billing for cloud services.  However, 
timekeeping in a VM context is very difficult.  Historically, timekeeping
is accomplished using software clocks based on hardware oscillators, and
keeping the software clocks synchronized to an online
reference clock.  However, the synchronization algorithms used rely on
computing delays between timestamped packets, and the delay computation
turns out to be extremely unstable in a virtualized context, due mostly
to added latency during packet processing.

This paper does not really introduce a novel idea, but rather applies
an existing alternative solution for timekeeping to the virtualized context,
and then performs some rigorous experimentation to show that this solution
is better-suited.

In the hardware, there are a few oscillators that may be used as sources
for timekeeping.  Most systems today make use of the \emph{Time Stamp Counter}
(TSC), a high-resolution counter based on CPU cycles, and the
\emph{High Precision Event Timer} (HPET), which (ironically enough) has a
lower resolution than the TSC, and is slower to access as well.  The authors
mention hardware counters only because the technique currently used for timekeeping
in virtualized systems, Xen's Clocksource algorithm, relies on these two counters,
using the HPET for low-resolution ``ticks'', but interpolating between them using
the TSC.  However, this TSC data must be carefully adjusted before use, because
it is subject to drift based on a variety of other variables.

The focus of the paper is more on timekeeping synchronization algorithms.
The most popular contender in this arena is the \emph{Network Time Protocol} (NTP).
NTP is a feedback algorithm, which works by collecting and timestamping a series
of packets from a networked reference clock, computing clock error using
the reference timestamp-local timestamp pairings as data points, 
and updating the system clock accordingly.  Xen Clocksource makes use of NTP,
although it must be adapted to fit the virtualized context.  There are two
possibilities for this:  in Dependent Clocksource, \emph{dom0} runs an NTP
clock, and all guests running in \emph{domU} sync periodically to this NTP
clock and use Clocksource to interpolate between syncs.  Alternatively, there
is an Independent Clocksource paradigm in which each \emph{domU} guest runs their own
NTP clock.  The paper conducts evaluation of both paradigms, and argues
that neither is well-suited to the virtualized context.  Dependent Clocksource
exhibits cyclical sawtoothed drift because the Clocksource algorithm in
\emph{domU} does not have direct access to the TSC, and so cannot adjust
the NTP clock data intelligently.  Independent Clocksource is extremely
inefficient, but even worse, additional latencies incurred by contention for
network I/O tend to push the Clocksource feedback algorithm into instability,
so that large error can accumulate even across NTP synchronizations.

The authors instead endorse an alternative sync algorithm, \emph{RADClock}
(Robust Absolute and Difference Clock).  RADClock is a stateless, feed-forward
algorithm.  Like NTP, RADClock timestamps reference packets, but does so using
a ``raw'' clock based on hardware counters.  The raw clock's error is estimated
as before, but rather than using this data to update the clock (which would make
RADClock a feedback algorithm), the error is instead stored in memory, along
with some other clock parameters for the period of the counter and an offset
for the timescale being used.  When the clock is read, the time is adjusted
as
\[
	C(t) = Raw(t) * (ctr~period) + (offset) - error(t)
\]
RADClock adjusts equally well to either the dependent or equally dependent paradigm,
since no interpolation is needed to maintain the clock between syncs, as long
as all dependent clocks can look up accurate and up-to-date parameters.

The paper proceeds to rigorously evaluate the performance of Xen Clocksource
using NTP against RADClock, both in dependent and independent modes of
operation, and in a variety of different system conditions.  The conclusion
seems to be that RADClock is generally unaffected by the errors and
drift that plague clocksource.  This is especially clear in a series of
experiments conducted involving VM migrations, where ``migration shock''
pushes the Independent Clocksource clock into a state of instability that
takes nearly 20 minutes to correct, while dependent RADClock converges
instantly.  Furthermore, the RADClock algorithm works equally well
regardless of the choice of underlying hardware counter, and is much
less affected by high contention for the network and system resources,
making it a better solution for the problem of timekeeping in the
context of virtualization.

\subsection{Paper 4 - TBD}
\label{sec:summaries/tbd}

\section{Paper Relationships}
\label{sec:relationships}

\section{Conclusion}
\label{sec:conclusion}

\begin{thebibliography}{10}

{\footnotesize

\bibitem{ref:xen}
	P. Barham, B. Dragovic, K. Fraser, S. Hand, T. Harris, A. Ho, R. Neugebauer,
	I. Pratt, A. Warfield.  Xen and the Art of Virtualization.  In {\it Proc. of
	the 19\textsuperscript{th} ACM Symp. on Operating Systems Principles}, 2003.

\bibitem{ref:turtles}
	M. Ben-Yehuda, M. Day, Z. Dubitzky, M. Factor, N. Har'El, A. Gordon, A. Liguori,
	O. Wasserman and B. Yassour.  The Turtles Project: Design and Implementation
	of Nested Virtualization.  In {\it Proc. of the 9\textsuperscript{th}
   USENIX Symp. on Operating Systems Design \& Implementation}, 2010.

\bibitem{ref:mclock}
	A. Gulati, A. Merchant, and P. Varman.  mClock: Handling Throughput Variability
	for Hypervisor IO Scheduling.  In {\it Proc. of the 9\textsuperscript{th}
   USENIX Symp. on Operating Systems Design \& Implementation}, 2010.

\bibitem{ref:time}
	T. Broomhead, L. Cremean, J. Ridoux, and D. Veitch.  Virtualize Everything but
	Time.  In {\it Proc. of the 9\textsuperscript{th}
   USENIX Symp. on Operating Systems Design \& Implementation}, 2010.

}

\end{thebibliography}

\end{multicols*}

\end{document}

### 2011-12-12 paper.tex
greg's stuff

@@
-\documentclass[10pt]{article}
+\documentclass[letterpaper, twocolumn]{article}
 \pagestyle{plain}
@@
 \usepackage{url}
-
-\begin{document}
+\usepackage{graphicx}
 
@@
 
+\begin{document}
+
+\twocolumn[
+\begin{@twocolumnfalse}
 \maketitle
@@
 
-\begin{multicols*}{2}
+\end{@twocolumnfalse}
+]
 
@@
 sorts of techniques which are utilized today.  Section \ref{sec:summaries}
-presents summaries of three recent virtualization papers from the
-9\textsuperscript{th} USENIX Symposium on Operating Systems Design
-and Implementation, as well as a fourth from (TBD!).  Section
-\ref{sec:relationships} analyses these papers in terms of how they
+presents summaries of a foundation virtualization paper from the
+5\textsuperscript{th} USENIX Symposium on Operating Systems Design
+and Implementation, as well as three recent papers from the virtualization
+session of the same Symposium in 2010.  Section \ref{sec:relns} attempts
+to analyze and classify these papers in terms of how they
 relate to each other and how they approach the larger problem space.
@@
 
+\subsection{Scale and Performance in the Denali Isolation Kernel}
+\label{sec:summaries/denali}
+
 \subsection{The Turtles Project: Design and Implementation of Nested Virtualization}
@@
 exist logically, and in reality they are all multiplexed to be running directly
-on top of $L_0$.  All traps generated at $L_2$ go immediately to $L_0$, which
+on top of $L_0$ (see Figure \ref{fig:turtles-multiplex}).  All traps generated at
+$L_2$ go immediately to $L_0$, which
 may handle them itself, or forward them back to $L_1$ for handling if hardware
 interaction is not required.
+
+\begin{figure*}[t]
+	\begin{center}
+		\includegraphics[width=4in]{images/turtles.png}
+	\end{center}
+	\caption{The Turtles hypervisor maintains logical nested virtualization, but
+		in reality multiplexes all levels of nesting into $L_1$. (Figure courtesy
+		of \cite{ref:turtles})}
+	\label{fig:turtles-multiplex}
+\end{figure*}
 
@@
 
-The mClock algorithm works by assigning numerical tags to each queued request.
+\begin{figure}[t]
+	\begin{center}
+		\includegraphics[width=2.5in]{images/mclock.png}
+	\end{center}
+	\caption{The mClock algorithm works by tagging requests with a tag for
+		reservations, weights, and reservations, and then executing requests
+		in order of their tags.  (Figure courtest of \cite{ref:mclock})}
+	\label{fig:mclock-alg}
+\end{figure}
+
+The mClock algorithm (see Figure \ref{fig:mclock-alg}) works by assigning numerical tags to each
+queued request.
 Essentially, tags for VM $i$ with proportional weight $w_i$ will be spaced by $1/w_i$.
@@
 seems to be that RADClock is generally unaffected by the errors and
-drift that plague clocksource.  This is especially clear in a series of
+drift that plague clocksource (see Figure \ref{fig:time-eval}).
+This is especially clear in a series of
 experiments conducted involving VM migrations, where ``migration shock''
@@
 
-\subsection{Paper 4 - TBD}
-\label{sec:summaries/tbd}
+\begin{figure*}[t]
+	\caption{The RADClock algorithm consistently outperforms ntpd clocksource;
+		here, the drift in clocksource is simply due to feedback from increased
+		latency in the network. (Figure courtesy of \cite{ref:time})}
+	\label{fig:time-eval}
+	\begin{center}
+		\includegraphics[width=6in]{images/time.png}
+	\end{center}
+\end{figure*}
 
 \section{Paper Relationships}
-\label{sec:relationships}
+\label{sec:relns}
+
+As we surveyed this series of papers, we noticed that the literature
+in this area seems to neatly classify itself by \emph{motivation}.
+Many papers on virtualization come from the realm of academia, and these
+are mostly characterized by a focus on clean abstractions and solutions
+with good theoretical properties, such as fairness and strong isolation.
+Many more papers, however, originate from corporate research at places
+like Microsoft and HP Labs.  These papers seem much more interested
+in providing fast, scalable solutions using commodity hardware and
+software.  In our papers, we found a series of tensions which seemed
+to be expressing this fundamental question of motivation.  We will
+now turn our attention to these tensions, keeping in mind that the
+Turtles and mClock papers share corporate authorship, while the
+Denali and Time papers have originated in the realm of academia.
+
+\subsection{Performance and Fairness}
+\label{sec:relns/perf-fair}
+
+The most obvious tension which surfaced in the literature was a question
+of priority between performance and fairness.  Generally speaking, there is a trend
+in corporate papers to be heavily concerned with speed and efficiency, while
+academically-minded papers seem more focused on strong properties of isolation
+and guarantees of fairness between VMs.  This should not be particularly surprising,
+given the authorship; more efficient virtualization seems to be an easily reachable
+step toward greater profit.  It is worth noting, however, that in the more recent
+literature corporate papers seem to be taking note of the importance of fairness;
+we will discuss this further in Section \ref{sec:relns/scale}.
+
+Our papers are clearly divided by this metric:  Turtles is all about efficiency
+and performance, often at the cost of fairness.  The problem seems
+especially hard here because of the nested context.  It seems perfectly logical that if
+a guest OS running in $L_2$ should only receive a fraction $\alpha$ of the resources allocated
+to its host in $L_1$, which in turn should only receive a percentage $\beta$ of the machine's
+total resources, then these two fractions should compound and our guest should end up with
+$\alpha\beta$ of the total capacity available.  However, in reality Turtles multiplexes
+all guests at all levels onto $L_0$, so in many ways the guest OS and its hypervisor will
+be treated as siblings rather than a parent and child.  Multi-level device assignment
+for Port I/O and Memory-mapped I/O is particularly agnostic to these concerns, because
+it works exactly the same way as single-level device assignment, and so doesn't care whether
+a guest runs at $L_1$ or $L_{100}$.  Furthermore, no experimentation at all is done to analyze
+the effects of this model on fairness, and the authors even make the statement that this
+is a secondary concern, and so they are content with a sort of best-effort.
+
+On the other hand, Denali and mClock are all about prioritizing fairness.  The foundational
+goal of the Denali isolation kernel is to ensure strong isolation between clients, so that
+the behavior of any particular VM cannot in any way affect another.  And the whole point of
+the mClock scheduling algorithm is to ensure that limits, reservations, and weights
+are met fairly between VMs contending for a resource.  It should be noted that neither
+of these papers particularly eschew performance as a worthy goal; rather, the view
+here seems more to be that fast performance for a particular guest doesn't really
+mean much if a malicious (or even just selfish neighbor) can somehow choke that VM
+out and make it impossible to make progress.
+
+The Time paper is somewhat agnostic to concerns of performance and fairness anyway,
+since it deals more with the abstractions presented to client applications.  We
+will turn to this next.
+
+\subsection{Paravirtualization and Full Virtualization}
+\label{sec:relns/para-full}
+
+The second tension seems to arise between the paradigms of \emph{paravirtualization}
+and \emph{full virtualization}, which were introduced in Section \ref{sec:background}.
+Corporately-funded literature seems to have an eye more toward using as much
+commodity software as possible here, for the sake of lower adoption costs, while
+academic papers seem more willing to consider modifying client software in
+appropriate situations.
+
+The split here is pretty clear as well: Denali, Xen, and Time all consider
+paravirtualization to be beneficial and even necessary in the right situations.
+In Denali, the isolation kernel presents a modified and simplified hardware abstraction
+to its clients, which cleverly allows for some big wins in efficiency and performance
+as a side effect.  Xen claims that paravirtualization is necessary to ensure fairness
+and prevent starvation in process and I/O scheduling; one could argue, however, that
+this is no longer a concern due to the advent of newer virtualization-aware hardware
+(see Section \ref{sec:relns/hard-emul}).  The Time paper goes a step further and claims that
+even with the new hardware, making the guest OS aware of the virtualized context it
+runs in is necessary to ensure the correctness of certain time-dependent applications!
+
+In contrast, the Turtles project argues for full virtualization and unmodified guest
+OSes and hypervisors.  In the paper, the argument for this goes back to the
+motivating question of efficiency, though it was unclear to us why paravirtualization
+would have any negative impact here.  But it seems also that for any sort of large-scale
+commercial deployment of server virtualization technology, the ability to use commodity
+guests would be crucial in order to make adoption easy, and we consider this to also
+be a likely factor influencing the design choice.
+
+The mClock paper doesn't really fit into these two classifications, since it mostly
+deals with the interactions between the hypervisor and its devices, rather than the hypervisor
+and its guests.  However, the paper does mention that one of the reasons this disk
+scheduling problem is so difficult is because guest OSes have their own algorithms
+for efficient disk scheduling, which generally don't play nice with each other.  If it
+were possible to modify the guest OS to disable these algorithms, so that the hypervisor
+could handle all disk scheduling decision, the difficulty of this problem could likely
+be reduced.
+
+\subsection{Hardware Support and Software Emulation}
+\label{sec:relns/hard-emul}
+
+The ``tension'' between hardware support for virtualization and the use of software
+emulation is perhaps better just described as a paradigm shift.  It seems
+that over the last decade virtualization-friendly hardware has become commonplace
+and inexpensive.  When the Xen and Denali papers were written, there was no real
+alternative, and so everything had to be done through slow, inefficient software emulation.
+But when hardware manufacturers began to produce virtualization-enabled chipsets
+like the x86, this problem essentially disappeared overnight, and it has largely
+remained out-of-scope for most of the intervening decade.
+
+However, the question of software emulation may need to be re-examined as virtualization
+research moves foward.  While hardware-enabled approaches are clearly much faster
+than any attempt to do all the work in software, it seems that it just may not be possible to do
+everything at the hardware level.  The Turtles project is a perfect example of this - there
+is no hardware support for nested levels of virtualization, and so a lot of cleverness
+is required to be able to compensate in the software without an unacceptable performance
+penalty.  We have essentially been sent back to where single-level virtualization
+solutions were a decade ago.  Some of the future work mentioned in this paper involved
+examining ways to provide architectural support for nested virtualization, but although
+we are somewhat skeptical that this will prove fruitful.  Attempting to provide
+support for an unbounded number of levels of nesting with finite hardware does
+not seem like it would be particularly successful.  It will be interesting to
+see what direction this sort of work takes in the future, and whether hardware
+designers will rise to the challenge of nested virtualization.
+
+\subsection{Scalability and the Cloud}
+\label{sec:relns/scale}
+
+The final tension we wish to consider is the question of scalability.  We have observed
+that there is a commercial interest in developing super-scalable server virtualization,
+largely because it has found such a unique niche with the market growth of cloud computing
+and infrastructure-as-a-service (IAAS).  While academic research is not uninterested in the question
+of scale or the cloud use case, we do not feel that they are as much of a driving factor in this realm.
+
+We can observe this trend in the papers we have been discussing.  The Turtles paper presents
+the motivation for nested virtualization using the argument of its suitability for
+infrastructure-as-a-service, as the two are essentially made for each other.  The mClock paper,
+although it does not particularly market itself
+for scalability, is motivated by the cloud-enabled scenario of many virtualized clients
+needing to schedule the use of a shared network disk.  Furthermore, the guarantees mClock
+makes about fairness, reservations, and proportional share are especially important
+in this scenario, as they allow IAAS providers to keep service-level agreements with
+their clients.  We find this particularly interesting, as it seems to indicate
+that in the cloud, fairness and performance may find themselves on equal footing in
+terms of importance (see Section \ref{sec:relns/perf-fair}).
+
+Interestingly, the Denali paper also has a particular eye to scalability, and a large
+portion of the evaluations section is devoted to demonstrating that the isolation kernel
+performs well even when supporting over 10,000 VMs.  However, even in Denali, scalability
+is just one goal amongst many, and the paper does not really consider applications
+to cloud computing, being several years too early.
+
+The Time paper does not really consider the question of scalability at all, being
+somwhat orthogonal to the problem of accurate timekeeping on a single VM.  The paper
+does point out that there are scalability benefits to the dependent RADClock algorithm,
+as each VM no longer needs to send out synchronization packets on the network, but can
+depend on the hypervisor to do this for all VMs that it hosts.  However, this is 
+in many ways just an added bonus, and is not really a motivational argument for the paper.
 
@@
 \label{sec:conclusion}
+
+Having examined the recent literature along these four axes of tension, it is our belief
+that a clear trend has emerged in virtualization.  We believe that as cloud computing and
+IAAS continue to take off, research in this area will continue to trend toward what
+is both popular and financially successful.  We expect that the majority of virtualization
+research will continue in the direction of full virtualization over paravirtualization,
+and that the cloud-computing/IAAS scenario will become the driving use case.
+However, we think that as a result of this, fairness and performance will no longer be
+competing ends, but will in fact become equal players.  This will be necessary in order
+for cloud providers to meet their SLAs to their consumers, and we believe fair-scheduling
+and resource-isolation research will move from its place in the shadows to the forefront
+of both corporate and academic research.
+
+As to the question of hardware support versus software emulation, we believe that as
+new directions like nested virtualization are explored, some sort of hybrid approach
+will continue to be needed.  However, we do not expect the status quo of fast hardware-enabled
+virtualization to be shaken up in the general case; it seems that performance is too critical
+in the driving cloud-based use cases, and we would be surprised to see nested virtualization
+(or any other software-hybrid approach) gaining momentum here anytime soon.
+
+Cynical though it may be, we feel that a lot of the interesting problems in virtualization
+have already been solved, at least from an academic perspective.  So much of the recent
+literature seems to be in the direction of just making things more efficient and more
+reliable for the sake of the commercialized cloud.  We have stopped asking the question,
+``can we do crazy thing $x$?'', and now are only asking ``how much faster can we make $x$ work?''
+and ``how can we make $x$ work with fewer resources?''
+This is certainly an important part of the research process for any area, but it is ostensibly
+less exciting than the ground-breaking work that was going on earlier in the decade.
+
+Of course, we cannot say with any certainty that there is not some new, groundbreaking discovery
+just waiting to be made around the corner.  Such a discovery could revitalize the field
+and take virtualization off into all sorts of new and exciting directions.  But we do feel that
+the way the problem has been framed makes it unlikely that such a new discovery could shift
+the focus away from cloud-based server virtualization.  With so much focus on making the problem
+fit into the market niche, such a new discovery would have to be truly monumental indeed -
+monumental enough that no one could possibly predict it.
 
@@
 	P. Barham, B. Dragovic, K. Fraser, S. Hand, T. Harris, A. Ho, R. Neugebauer,
-	I. Pratt, A. Warfield.  Xen and the Art of Virtualization.  In {\it Proc. of
+	I. Pratt, and A. Warfield.  Xen and the Art of Virtualization.  In {\it Proc. of
 	the 19\textsuperscript{th} ACM Symp. on Operating Systems Principles}, 2003.
+
+\bibitem{ref:denali}
+	A. Whitaker, M. Shaw, and S. Gribble.  Scale and Performance in the Denali Isolation
+	Kernel.  In {\it Proc. of the 5\textsuperscript{th} Symp. on Operating Systems
+	Desgin \& Implementation}, 2002.
 
@@
 
-\end{multicols*}
-
 \end{document}

### 2011-12-12 paper.tex
denali summary added

@@
 \label{sec:summaries/denali}
+
+Three years after VMWare's appearance, Denali sought to provide a virtualization
+solution design with two goals in mind, scalability and security. Denali's authors
+envisioned a separation between the management of physical hardware and the management
+of software services, transforming how internet services could be deployed. Such a system 
+would allow untrusted software to run on third party infrastructure, support the distribution 
+of dynamic content in delivery systems, or provide virtual hosting infrastructure.
+
+In order to meet this vision the Denali project introduced \emph{para-virtualization} in the form 
+of isolation kernels, adopted complete isolation of guest virtual machines, utilizes lightweight
+virtual machines to support over 10,000 VMs on commodity machines. However, Denali diverges from many 
+related projects by sacrificing backwards compatibility for its primary objectives, but its authors argue
+tha this design trade off is worth the benefits and suggest that legacy operating systems could be
+ported to Denali.
+
+At the high level, Denali provides an isolation kernel, a small kernel operating system, running
+on top of x86 hardware. The kernel is designed to host many untrusted applications which are isolated
+and require little data sharing. Denali multiplexes between the guest operating systems and isolates them
+using private virtualized namespaces. Some special privileges, 
+like dowloading a remote VM or loading initial disk images, are handled by a supervisor VM.
+Since the x86 ISA includes instructioins whose behavior differs
+in user and kernel mode, Denali provides a subset of the x86 ISA as well as two virtual instructions.
+First, idle-with-timeout lets VMs halt their execution for a bounded amount of physical time or until
+an interrupt arrives for the VM in return for a higher priority, 
+and secondly Denali provides a virtual instruction which VMs can use to terminate.
+
+Denali's virtual architecture and implementation adhere to a separation between mechanism and policy 
+so changes can be easily made. CPU virtualization holds a policy split into two elements, 
+a \emph{gatekeeper policy} to choose the subset of active VMs to admit into the system and 
+the \emph{scheduler policy} to switch between active VMs. For simplicity, the authors use FIFO
+and round robin in the descirbed implementation. 
+% maybe new paragraph? and expand the memory virtualization stuff?
+Denali's memory virtualization aims to keep a VM's working set in memory, so physical memory 
+is redistributed from inactive to active VMs, and swap regions for each VM are used to page its memory.
+% maybe new paragraph?
+Denali also provides a virtual ethernet NIC for each VM with queues for incoming and outgoing packets.
+Incoming packets for each VM are accumulated in the appropriate queues and are later process once the
+corresponding VM is given focus by the scheduler, whereas outgoing packets are processed in a round robin manner
+from outgoing queues. The fact that Denali's TCP/IP stack runs at the user level has its drawbacks, 
+since all packets have to cross the user/kernel boundary. However, the packet dispatch evaluation 
+suggested that cost of network virtualization was low and that the the driver was responsible for most of the
+costs of packet processing.
+% new p?
+% The Denali isolation kernel delivers interrupts in batches to VMs
+% what else to say about interrupts
+% Denali provides simple fixed size virtual disks to VMs and basic devices (keyboard, console, etc) not really important?
+
+The Denali paper provides a comprehensive evaluation of the Denali isolation kernel running
+the Ilwaco operating system with basic metrics like switching times between VMs, swap disk
+microbenchmarks, and packet dispatch latency, and TCP and HTTP throughput. Some of the most
+interesting evaluations were conducted to examine the scalability and performance of Denali.
+The authors found that batched interrupts provided a performance gain of 30\% when the number
+of VMs exceeded 800, and that using idle-with-timeout in Denali lead to higher performance as
+the number of VMs increased.
+
+However, Denali is still not free from bottlenecks, and the authors discuss two major challenges to
+Denali's performanc and scalability. First, Denali tries to reduce the amount of per-VM kernel data
+that it has too manage by no keeping information for inactive VMs, but it still requires about 81MB
+of data for 10,000 virtual machines. The authors suggest that with the growing size of memory,
+this bottleneck might be acceptable. Secondly, the paging behavior of the guest is crucial to overall performance,
+and while Denali does its best to keep VMs' working sets in memory, a misbehaving guest can severly damage the
+performance of the system. One possible solution to this problem might lie with memory management 
+and garbage collection, since ensuring good behavior and providing good locality could ameliorate this problem.
+
+% transition to present day?
 

### 2011-12-12 paper.tex
typos

@@
 
-Three years after VMWare's appearance, Denali sought to provide a virtualization
+Three years after VMWare's appearance, Denali \cite{ref:denali} sought to provide a virtualization
 solution design with two goals in mind, scalability and security. Denali's authors
@@
 like dowloading a remote VM or loading initial disk images, are handled by a supervisor VM.
-Since the x86 ISA includes instructioins whose behavior differs
+Since the x86 ISA includes instructions whose behavior differs
 in user and kernel mode, Denali provides a subset of the x86 ISA as well as two virtual instructions.

### 2011-12-13 paper.tex
greg's edits

--- 2011-12-12 paper.tex
@@
 
-The concept of virtualization has been around for a long time, reaching back
+The idea of virtualization has been around for a long time, reaching back
 to the 1960s and the introduction of virtual memory in timeshared machines.
@@
 on the same hardware.  Companies can now spend less on metal, can seamlessly and
-easily migrate servers from one machine to another, and can even vary their fleet
+easily migrate servers from one machine to another, and can even adjust their fleet
 size on the fly in response to fluctuations in server traffic by spinning up new
@@
 community as well, ranging from new solutions to problems of isolation and protection
-to novel new algorithms for scheduling and resource allocation.
+to novel algorithms for scheduling and resource allocation.
 
-Despite the flurry of interest in this area, there are still many interesting
+Despite the recent flurry of activity in this area, there are still many interesting
 and difficult problems to be solved.  Sharing finite resources
 fairly and safely amongst many competing processes has never been easy, but
-the popularity of virtualization in the cloud makes scalability and efficiency
+the popularity of virtualization in the Cloud makes scalability and efficiency
 bigger concerns than ever before.  Furthermore, the commercialization of
@@
 sorts of techniques which are utilized today.  Section \ref{sec:summaries}
-presents summaries of a foundation virtualization paper from the
+presents summaries of a foundational virtualization paper from the
 5\textsuperscript{th} USENIX Symposium on Operating Systems Design
@@
 
-The concept of virtualization reaches back to the 1960s, when developers
+Virtualization got it's start back in the 1960s, when developers
 of large time-sharing systems sought ways to give each user the illusion that
@@
 in virtualization has sparked significantly, especially in conjunction
-with the rise of cloud computing and infrastructure-as-a-service.
+with the rise of Cloud computing and Infrastructure-as-a-service.
 
@@
 that it alone has sole access to the hardware, although in a \emph{paravirtualized}
-VMs may be modified to be aware of the fact that they are virtualized.  This
+system VMs may be modified to be aware of the fact that they are virtualized.  This
 sharing of resources is possible because access to the hardware is regulated by
@@
 domain (in Xen, this is called \emph{dom0}), while other OSes running in VMs are 
-referred to as \emph{guests} (in Xen, all guests run in \emph{domU}).  Generally, it is
+referred to as \emph{guests} (in Xen, all guests run in \emph{domU}).  It is
 crucial that all guests are isolated from one another, so that the actions of one guest
 cannot adversely affect the performance of any other.  This is a difficult property
-to provide in a system where all guests are in contention for hardware resources.
+to guarantee in a system where all guests are in contention for limited hardware resources.
 
@@
 the hypervisor, and so most device accesses by a guest VM resulted in multiple traps
-into the hypervisor.  In the past decade or so, chipsets have become more friendly,
+into the hypervisor (and therefore multiple crossings of protection boundaries).
+In the past decade or so, chipsets have become more virtualization-friendly,
 and most architectures (notably the x86) now offer hardware-level support for
@@
 \label{sec:summaries/denali}
+
+Three years after VMWare's appearance, Denali \cite{ref:denali} sought to provide a virtualization
+solution design with two goals in mind, scalability and security. Denali's authors
+envisioned a separation between the management of physical hardware and the management
+of software services, transforming how internet services could be deployed. Such a system 
+would allow untrusted software to run on third party infrastructure, support the distribution 
+of dynamic content in delivery systems, or provide virtual hosting infrastructure.
+
+In order to meet this vision, the Denali project introduced \emph{paravirtualization} by providing
+a simplified and pared-down emulation of the hardware for VMs to interface with.  Denali also
+adopted complete isolation of guest VMs, and made VMs extremely lightweight, such that the system can
+support over 10,000 VMs on commodity machines. Denali diverges from many 
+related projects by sacrificing backwards compatibility for its primary objectives,
+but its designers argue that this design tradeoff is well worth the benefits.
+
+At the high level, Denali provides an isolation kernel (a small kernel operating system) running
+on top of x86 hardware. The kernel is designed to host many untrusted applications which are isolated
+and require little-to-no data sharing. Denali multiplexes between the guest operating systems and isolates them
+using private virtualized namespaces. Some special privileges, like dowloading a remote VM or loading initial
+disk images, are handled by a supervisor host VM.
+
+The isolation kernel exports a simpler interface to guest VMs than would be required by a
+traditional operating system.  Since the x86 ISA includes instructions whose behavior differs
+in user and kernel mode, these are excluded from Denali's instruction set,
+while two additional virtual instructions are added.
+First, \verb+idle-with-timeout+ lets VMs halt their execution
+for a bounded amount of physical time, or until an interrupt arrives for the VM,
+which allows the system to avoid needlessly scheduling sleeping VMs.
+Second, Denali provides a virtual instruction which a VM can use to self-terminate.
+
+Denali's virtual architecture and implementation adhere to a separation between mechanism and policy 
+so changes can be easily made. CPU virtualization holds a policy split into two elements, 
+a \emph{gatekeeper policy} to choose the subset of active VMs to admit into the system, and 
+a \emph{scheduler policy} to switch between active VMs. For simplicity, the authors use FIFO
+and round robin in the descirbed implementation, although other scheduling policies could
+be implemented.  The system provides a non-traditional interrupt model, where virtual interrupts
+for a VM are batched and delivered asynchronously when the VM is next scheduled.
+
+Denali's memory virtualization aims to keep a VM's working set in memory, so physical memory 
+is redistributed from inactive to active VMs, and swap regions for each VM are used to page its memory.
+The isolation kernel itself is kept pinned in memory at all times.
+
+Denali also provides some limited emulated hardware.  In particular, a virtual ethernet NIC is 
+provided for each VM, with queues for incoming and outgoing packets.
+Incoming packets for each VM are accumulated in the appropriate queues and are later processed once the
+corresponding VM is given focus by the scheduler.  Outgoing packets are processed in a
+round-robin manner from outgoing queues. Denali's TCP/IP stack runs at the user level, which
+has its drawbacks, since all packets have to cross the user/kernel boundary. However, the paper contains
+a packet dispatch evaluation, which suggests that cost of network virtualization is low and that the the
+driver itself is responsible for the majority of the costs of packet processing.
+
+The Denali paper provides a comprehensive evaluation of the isolation kernel running
+the Ilwaco operating system, with basic metrics like switching times between VMs, swap disk
+microbenchmarks, and packet dispatch latency, and TCP and HTTP throughput. Some of the most
+interesting evaluations were conducted to examine the scalability and performance of Denali.
+The authors found that batched interrupts provided a performance gain of 30\% when the number
+of VMs exceeded 800, and that using idle-with-timeout in Denali lead to higher performance as
+the number of VMs increased.
+
+However, Denali is certainly not free from bottlenecks, and the authors discuss two major challenges to
+Denali's performance and scalability. First, Denali tries to reduce the amount of per-VM kernel data
+that it has to manage by no keeping information for inactive VMs, but it still requires about 81MB
+of data for 10,000 virtual machines. The authors suggest that with the growing size of memory,
+this bottleneck might be acceptable. Secondly, the paging behavior of the guest is crucial to overall performance,
+and while Denali does its best to keep VMs' working sets in memory, a misbehaving guest can severly damage the
+performance of the system.  The authors claim that this problem may be solvable with good memory management 
+and garbage collection techniques, which could increase locality.
 
@@
 we will call $L_0$, at the most privileged level of execution.  Now, suppose
-we have a hypervisor running above $L_0$ at level $L_1$, and that hypervisor
-is running a guest OS at level $L_2$.  Everything running at $L_1$ and $L_2$
-has to run at the same privilege level, and only $L_0$ is privileged enough
+we have a hypervisor running above $L_0$ at level $L_1$, and further suppose 
+that hypervisor is running a guest OS at level $L_2$.  Everything running at $L_1$
+and $L_2$ has to run at the same privilege level, and only $L_0$ is privileged enough
 to interact with the hardware.  Essentially, levels $L_1$ through $L_n$ only
-exist logically, and in reality they are all multiplexed to be running directly
+exist logically; in reality they are all multiplexed to be running directly
 on top of $L_0$ (see Figure \ref{fig:turtles-multiplex}).  All traps generated at
-$L_2$ go immediately to $L_0$, which
-may handle them itself, or forward them back to $L_1$ for handling if hardware
-interaction is not required.
+$L_2$ go immediately to $L_0$, which may handle them itself, or forward them
+back to $L_1$ for handling if hardware interaction is not required.
 
@@
 translation tables by emulating them in software is slow, particularly on updates,
-for the reasons described above.  The paper's novel solution here is to use
+for reasons described previously.  The paper's novel solution here is to use
 \emph{multidimensional page tables}.  Essentially, each level keeps its own
@@
 it to update its shadow table with the $L_2$-to-$L_1$ translation, but then applies
-its tranlation for $L_1$ addresses to update the hardware table with the correct
+its tranlation for $L_1$ addresses to update the hardware-level table with the correct
 $L_2$-to-$L_0$ translation before returning.  In this way, the hardware table at $L_0$
@@
 also employs ``huge pages'' and Virtual Processor ID tags on page table entries
-to ensure that page tables remain small and fast.
+to ensure that page tables remain small and fast, and to prevent excessive TLB
+flushing on context-switches.
 
@@
 and Network Port I/O, address translation is not an issue, and so here techniques for
-single-level device assignment are used without any penalty.
+single-level device assignment are used without any efficiency penalty, although
+this does raise questions of fairness.
 
@@
 seemed to indicate that most of the additional overhead incurred by adding a second level
-came from cache pollution and the overuse of privileged instructions where it was not
+comes from cache pollution and the overuse of privileged instructions where it was not
 necessary.
@@
 out to be a very difficult problem even in the single-level case, particularly
-when applied to virtualization in the cloud, where networked devices are
+when applied to virtualization in the Cloud, where networked devices are
 often shared not just between VMs running on one host, but between many hosts
@@
 accesses may be reordered by the kernel to take advantage of spatial locality
-when moving the disk head.  These sorts of optimizations are often difficult
-for the hypervisor to multiplex without loss of efficiency.
+when moving the disk head.  These sorts of optimizations can make it difficult
+for the hypervisor to multiplex and schedule I/O requests efficiently.
 
-The mClock algorithm seeks to fairly schedule the device I/O of multiple guests,
-while maintaining each guest's limit (maximum throughput requirement), reservation
+The mClock algorithm seeks to fairly schedule the device I/O requests of multiple guests,
+while maintaining each guest's limit (maximum throughput cap), reservation
 (minimum requirement), and weight (proportional share of capacity relative to
@@
 The mClock algorithm (see Figure \ref{fig:mclock-alg}) works by assigning numerical tags to each
-queued request.
-Essentially, tags for VM $i$ with proportional weight $w_i$ will be spaced by $1/w_i$.
+queued request.  Essentially, tags for VM $i$ with proportional weight $w_i$ will be spaced by $1/w_i$.
 Requests will be executed in the order of their assigned tags, which ensures that
@@
 request $j$ is set as follows:
-
 \[
@@
 \]
-
 After the tagging, the request with the lowest tag will be scheduled.
@@
 a distributed storage system, which multiplexes VM requests fairly across multiple
-storate servers.  Finally, there's a heuristic for setting reservations for
+storage servers.  Finally, there's a heuristic for setting reservations for
 all VMs by default, so that no VM will be starved.
@@
 The algorithm is successful at delivering on its guarantees despite contention,
-which is important for ensuring VM isolation, and also has implications in the cloud,
-where meeting quality of service guarantees determine financial success.
+which is important for ensuring VM isolation, and also has implications in the Cloud,
+where the ability to meet quality-of-service guarantees has a high impact on
+financial success.
 
@@
 applications to everything from systems measurement and distributed database
-consistency to high-speed trading and billing for cloud services.  However, 
+consistency to high-speed trading and billing for Cloud services.  However, 
 timekeeping in a VM context is very difficult.  Historically, timekeeping
-is accomplished using software clocks based on hardware oscillators, and
-keeping the software clocks synchronized to an online
+is accomplished using software clocks based on hardware oscillators, which
+are kept synchronized to an online
 reference clock.  However, the synchronization algorithms used rely on
@@
 although it must be adapted to fit the virtualized context.  There are two
-possibilities for this:  in Dependent Clocksource, \emph{dom0} runs an NTP
+possibilities for this:
+\begin{itemize}
+\item in Dependent Clocksource, \emph{dom0} runs an NTP
 clock, and all guests running in \emph{domU} sync periodically to this NTP
-clock and use Clocksource to interpolate between syncs.  Alternatively, there
-is an Independent Clocksource paradigm in which each \emph{domU} guest runs their own
-NTP clock.  The paper conducts evaluation of both paradigms, and argues
+clock and use Clocksource to interpolate between syncs.  
+\item under the Independent Clocksource paradigm, each \emph{domU} guest runs their own
+NTP clock and send their own reference packets.
+\end{itemize}
+The paper conducts evaluation of both paradigms, and argues
 that neither is well-suited to the virtualized context.  Dependent Clocksource
@@
 Turtles and mClock papers share corporate authorship, while the
-Denali and Time papers have originated in the realm of academia.
+Denali, Xen, and ``Virtualize Everything but Time'' papers originate
+in the realm of academia.
 
@@
 and guarantees of fairness between VMs.  This should not be particularly surprising,
-given the authorship; more efficient virtualization seems to be an easily reachable
-step toward greater profit.  It is worth noting, however, that in the more recent
-literature corporate papers seem to be taking note of the importance of fairness;
-we will discuss this further in Section \ref{sec:relns/scale}.
+given the authorship; more efficient virtualization seems a logical step toward maximizing
+profit, which has a certain appeal in the corporate sphere.  It is worth noting, however,
+that in the more recent literature corporate papers seem to be taking note of the
+importance of fairness; we will discuss this further in Section \ref{sec:relns/scale}.
 
@@
 and performance, often at the cost of fairness.  The problem seems
-especially hard here because of the nested context.  It seems perfectly logical that if
+especially hard here when phrased in the nested context.  It seems perfectly logical that if
 a guest OS running in $L_2$ should only receive a fraction $\alpha$ of the resources allocated
@@
 all guests at all levels onto $L_0$, so in many ways the guest OS and its hypervisor will
-be treated as siblings rather than a parent and child.  Multi-level device assignment
-for Port I/O and Memory-mapped I/O is particularly agnostic to these concerns, because
+be treated equally, rather than as parent and child.  Multi-level device assignment
+for Port I/O and Memory-mapped I/O is particularly agnostic to fairness concerns, because
 it works exactly the same way as single-level device assignment, and so doesn't care whether
-a guest runs at $L_1$ or $L_{100}$.  Furthermore, no experimentation at all is done to analyze
-the effects of this model on fairness, and the authors even make the statement that this
-is a secondary concern, and so they are content with a sort of best-effort.
+a guest runs at $L_1$ or $L_{100}$ - all guests get equal treatment.  Furthermore, no
+experimentation at all is done to analyze the effects of this model on fairness;
+the authors even make the statement that this is a secondary concern,
+and so they are content with a sort of best-effort.
 
@@
 of these papers particularly eschew performance as a worthy goal; rather, the view
-here seems more to be that fast performance for a particular guest doesn't really
-mean much if a malicious (or even just selfish neighbor) can somehow choke that VM
+here seems more to be that fast performance for a particular VM doesn't really
+matter much if a malicious (or even just selfish) neighbor can somehow choke that VM
 out and make it impossible to make progress.
@@
 even with the new hardware, making the guest OS aware of the virtualized context it
-runs in is necessary to ensure the correctness of certain time-dependent applications!
+runs in is necessary to ensure the correctness of certain time-dependent applications,
+such as systems measurement, high-speed trading, and even video gaming.
 
@@
 OSes and hypervisors.  In the paper, the argument for this goes back to the
-motivating question of efficiency, though it was unclear to us why paravirtualization
+motivating goal of performance, though it was unclear to us why paravirtualization
 would have any negative impact here.  But it seems also that for any sort of large-scale
@@
 could handle all disk scheduling decision, the difficulty of this problem could likely
-be reduced.
+be reduced.  This seems to be an argument supportive of paravirtualization.
 
@@
 The ``tension'' between hardware support for virtualization and the use of software
-emulation is perhaps better just described as a paradigm shift.  It seems
+emulation is perhaps better described as a paradigm shift.  It seems
 that over the last decade virtualization-friendly hardware has become commonplace
@@
 is required to be able to compensate in the software without an unacceptable performance
-penalty.  We have essentially been sent back to where single-level virtualization
-solutions were a decade ago.  Some of the future work mentioned in this paper involved
-examining ways to provide architectural support for nested virtualization, but although
+penalty.  In some ways, this problem has essentially sent us back to where single-level virtualization
+solutions were a decade ago.  Some of the future work mentioned in the Turtles paper involved
+examining ways to provide architectural support for nested virtualization, but
 we are somewhat skeptical that this will prove fruitful.  Attempting to provide
@@
 see what direction this sort of work takes in the future, and whether hardware
-designers will rise to the challenge of nested virtualization.
+designers will rise to the challenges of nested virtualization.
 
@@
 
-The final tension we wish to consider is the question of scalability.  We have observed
+The final tension we wish to consider is the question of scalability and designing
+for the Cloud.  We have observed
 that there is a commercial interest in developing super-scalable server virtualization,
-largely because it has found such a unique niche with the market growth of cloud computing
-and infrastructure-as-a-service (IAAS).  While academic research is not uninterested in the question
-of scale or the cloud use case, we do not feel that they are as much of a driving factor in this realm.
+largely because it has found such a unique niche with the market growth of Cloud computing
+and Infrastructure-as-a-service (IAAS).  While academic research is not uninterested in the question
+of scale or the Cloud use case, we do not feel that they are as much of a driving factor in this realm.
 
-We can observe this trend in the papers we have been discussing.  The Turtles paper presents
+We see this trend in the papers we have been discussing as well.  The Turtles paper presents
 the motivation for nested virtualization using the argument of its suitability for
-infrastructure-as-a-service, as the two are essentially made for each other.  The mClock paper,
+Infrastructure-as-a-service, as the two are essentially made for each other.  The mClock paper,
 although it does not particularly market itself
-for scalability, is motivated by the cloud-enabled scenario of many virtualized clients
+for scalability, is motivated by the Cloud-enabled scenario of many virtualized clients
 needing to schedule the use of a shared network disk.  Furthermore, the guarantees mClock
@@
 their clients.  We find this particularly interesting, as it seems to indicate
-that in the cloud, fairness and performance may find themselves on equal footing in
-terms of importance (see Section \ref{sec:relns/perf-fair}).
+that in the Cloud, fairness and performance may find themselves on equal footing in
+terms of importance, despite what we have observed in Section \ref{sec:relns/perf-fair}.
 
@@
 is just one goal amongst many, and the paper does not really consider applications
-to cloud computing, being several years too early.
+to Cloud computing, being written several years before the Cloud really became prominent.
 
@@
 as each VM no longer needs to send out synchronization packets on the network, but can
-depend on the hypervisor to do this for all VMs that it hosts.  However, this is 
+depend on the hypervisor to do this once for all VMs that it hosts.  However, this is 
 in many ways just an added bonus, and is not really a motivational argument for the paper.
@@
 Having examined the recent literature along these four axes of tension, it is our belief
-that a clear trend has emerged in virtualization.  We believe that as cloud computing and
+that a clear trend has emerged in virtualization research.  We believe that as Cloud computing and
 IAAS continue to take off, research in this area will continue to trend toward what
@@
 research will continue in the direction of full virtualization over paravirtualization,
-and that the cloud-computing/IAAS scenario will become the driving use case.
-However, we think that as a result of this, fairness and performance will no longer be
-competing ends, but will in fact become equal players.  This will be necessary in order
-for cloud providers to meet their SLAs to their consumers, and we believe fair-scheduling
-and resource-isolation research will move from its place in the shadows to the forefront
-of both corporate and academic research.
+and that the Cloud-computing/IAAS scenario will become the driving use case.
+However, we think that as a result, fairness and performance will no longer be
+competing ends, but will in fact become equal goals in virtualized system design.
+This will be necessary in order for Cloud providers to meet their SLAs to their
+consumers, and we believe fair-scheduling and resource-isolation research will move
+from the shadows to the forefront of both corporate and academic research.
 
@@
 virtualization to be shaken up in the general case; it seems that performance is too critical
-in the driving cloud-based use cases, and we would be surprised to see nested virtualization
-(or any other software-hybrid approach) gaining momentum here anytime soon.
+in the driving Cloud-based use cases, and we would be surprised to see nested virtualization
+(or any other software-hybrid approach) gaining much momentum here anytime soon.
 
@@
 literature seems to be in the direction of just making things more efficient and more
-reliable for the sake of the commercialized cloud.  We have stopped asking the question,
+reliable for the sake of the commercialized Cloud.  We have stopped asking the question,
 ``can we do crazy thing $x$?'', and now are only asking ``how much faster can we make $x$ work?''
-and ``how can we make $x$ work with fewer resources?''
-This is certainly an important part of the research process for any area, but it is ostensibly
+and ``how can we make $x$ work with fewer resources?''  This is certainly an important part of
+the research process for any area, but it is ostensibly
 less exciting than the ground-breaking work that was going on earlier in the decade.
@@
 Of course, we cannot say with any certainty that there is not some new, groundbreaking discovery
-just waiting to be made around the corner.  Such a discovery could revitalize the field
+just waiting to be made and lurking around the corner.  Such a discovery could revitalize the field
 and take virtualization off into all sorts of new and exciting directions.  But we do feel that
 the way the problem has been framed makes it unlikely that such a new discovery could shift
-the focus away from cloud-based server virtualization.  With so much focus on making the problem
+the focus away from Cloud-based server virtualization.  With so much focus on making the problem
 fit into the market niche, such a new discovery would have to be truly monumental indeed -

### 2011-12-13 paper.tex
small edits

@@
 
-Virtualization got it's start back in the 1960s, when developers
+Virtualization techniques arose more than half a century ago in the early 1960s, when developers
 of large time-sharing systems sought ways to give each user the illusion that
@@
 Server virtualization arrived on the scene about a decade ago, with VMWare
-appearing in 1999 and Xen in 2003.  Since then, research and commercial interest
+appearing in 1999, Denali in 2002, and Xen in 2003.  Since then, research and commercial interest
 in virtualization has sparked significantly, especially in conjunction
@@
 physical hardware.  In a \emph{fully virtualized} system, each VM has the illusion
-that it alone has sole access to the hardware, although in a \emph{paravirtualized}
+that it alone has sole access to the hardware, although in a \emph{para-virtualized}
 system VMs may be modified to be aware of the fact that they are virtualized.  This
@@
 Three years after VMWare's appearance, Denali \cite{ref:denali} sought to provide a virtualization
-solution design with two goals in mind, scalability and security. Denali's authors
+solution with two goals in mind, scalability and security. Denali's authors
 envisioned a separation between the management of physical hardware and the management
@@
 would allow untrusted software to run on third party infrastructure, support the distribution 
-of dynamic content in delivery systems, or provide virtual hosting infrastructure.
+of dynamic content in delivery systems, or provide a foundation for virtual hosting.
 
-In order to meet this vision, the Denali project introduced \emph{paravirtualization} by providing
+In order to meet this vision, the Denali project introduced \emph{para-virtualization} by providing
 a simplified and pared-down emulation of the hardware for VMs to interface with.  Denali also
-adopted complete isolation of guest VMs, and made VMs extremely lightweight, such that the system can
-support over 10,000 VMs on commodity machines. Denali diverges from many 
+adopted complete isolation of guest VMs, and made VMs extremely lightweight, so that the system could
+support over 10,000 VMs on commodity machines. Denali diverged from many 
 related projects by sacrificing backwards compatibility for its primary objectives,
-but its designers argue that this design tradeoff is well worth the benefits.
+but its designers argued that this design tradeoff was well worth the benefits.
 
@@
 a \emph{scheduler policy} to switch between active VMs. For simplicity, the authors use FIFO
-and round robin in the descirbed implementation, although other scheduling policies could
+and round robin in the described implementation, although other scheduling policies could
 be implemented.  The system provides a non-traditional interrupt model, where virtual interrupts
@@
 a packet dispatch evaluation, which suggests that cost of network virtualization is low and that the the
-driver itself is responsible for the majority of the costs of packet processing.
+driver itself is responsible for the majority of packet processing costs.
 
@@
 \emph{multi-level device assignment}.  The Turtles hypervisor is able to
-run unmodified guest hypervisors and OSes with a performance penalty of
+run unmodified guest hypervisors and OSes with a performance penalty
 as low as 6-8\%.  The paper also presents what the authors claim is the
 first thorough analysis and evaluation of nested hypervisor performance,
-with an eye to the identification and reduction of performance bottlenecks.
+with an eye to the identification and reduction of performance bottlenecks. % not sure about ``with an eye to''
 
@@
 
-\begin{figure}[t]
-	\begin{center}
-		\includegraphics[width=2.5in]{images/mclock.png}
-	\end{center}
+\begin{figure}[ht!]
+	%\begin{center}
+		\includegraphics[width=2.65in]{images/mclock.png}
+	%\end{center}
 	\caption{The mClock algorithm works by tagging requests with a tag for
@@
 determined after the tagging:  if any VM $i$ has request $j$ such that the
-reservation tag $R(i, j)$ is less than the current time, than VM $i$ is currently
+reservation tag $R(i, j)$ is less than the current time, then VM $i$ is currently
 receiving less capacity than its reservation, and so the current phase is
@@
 and performance, often at the cost of fairness.  The problem seems
-especially hard here when phrased in the nested context.  It seems perfectly logical that if
+especially hard here when phrased in the nested context.  It seems perfectly logical that
 a guest OS running in $L_2$ should only receive a fraction $\alpha$ of the resources allocated
@@
 
-\subsection{Paravirtualization and Full Virtualization}
+\subsection{Para-virtualization and Full Virtualization}
 \label{sec:relns/para-full}
 
-The second tension seems to arise between the paradigms of \emph{paravirtualization}
+The second tension seems to arise between the paradigms of \emph{para-virtualization}
 and \emph{full virtualization}, which were introduced in Section \ref{sec:background}.
@@
 The split here is pretty clear as well: Denali, Xen, and Time all consider
-paravirtualization to be beneficial and even necessary in the right situations.
+para-virtualization to be beneficial and even necessary in the right situations.
 In Denali, the isolation kernel presents a modified and simplified hardware abstraction
 to its clients, which cleverly allows for some big wins in efficiency and performance
-as a side effect.  Xen claims that paravirtualization is necessary to ensure fairness
+as a side effect.  Xen claims that para-virtualization is necessary to ensure fairness
 and prevent starvation in process and I/O scheduling; one could argue, however, that
@@
 OSes and hypervisors.  In the paper, the argument for this goes back to the
-motivating goal of performance, though it was unclear to us why paravirtualization
+motivating goal of performance, though it was unclear to us why para-virtualization
 would have any negative impact here.  But it seems also that for any sort of large-scale
@@
 could handle all disk scheduling decision, the difficulty of this problem could likely
-be reduced.  This seems to be an argument supportive of paravirtualization.
+be reduced.  This seems to be an argument in favor of para-virtualization.
 
@@
 largely because it has found such a unique niche with the market growth of Cloud computing
-and Infrastructure-as-a-service (IAAS).  While academic research is not uninterested in the question
+and Infrastructure-as-a-service (IaaS).  While academic research is not uninterested in the question
 of scale or the Cloud use case, we do not feel that they are as much of a driving factor in this realm.
@@
 makes about fairness, reservations, and proportional share are especially important
-in this scenario, as they allow IAAS providers to keep service-level agreements with
+in this scenario, as they allow IaaS providers to keep service-level agreements with
 their clients.  We find this particularly interesting, as it seems to indicate
@@
 Interestingly, the Denali paper also has a particular eye to scalability, and a large
-portion of the evaluations section is devoted to demonstrating that the isolation kernel
+portion of the evaluation section is devoted to demonstrating that the isolation kernel
 performs well even when supporting over 10,000 VMs.  However, even in Denali, scalability
@@
 that a clear trend has emerged in virtualization research.  We believe that as Cloud computing and
-IAAS continue to take off, research in this area will continue to trend toward what
+IaaS continue to take off, research in this area will continue to trend toward what
 is both popular and financially successful.  We expect that the majority of virtualization
-research will continue in the direction of full virtualization over paravirtualization,
-and that the Cloud-computing/IAAS scenario will become the driving use case.
+research will continue in the direction of full virtualization over para-virtualization,
+and that the Cloud-computing/IaaS scenario will become the driving use case.
 However, we think that as a result, fairness and performance will no longer be
